---
title: 'Lab 9: Working with COVID-19 Reporting Data'
author: "Jarrod Daniels"
date: "3/25/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, results='hide'}
library(tidyverse)
library(lubridate)
```

```{r load data, results='hide'}
report_03_11_2020 <-   read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/03-11-2020.csv")) %>%
  rename(Country.Region = "Country/Region", Province.State = "Province/State")
```

Checking the table properties to make sure the data imported as we expected
```{r head}
head(report_03_11_2020)
```

```{r str}
str(report_03_11_2020)
```

Now create a plot of confirmed cases in the US as of March 11th
```{r confirmedCaseState, fig.height = 7.5}
report_03_11_2020 %>% 
  filter (Country.Region == "US") %>% 
  ggplot(aes(x = Confirmed,  y = reorder(Province.State, Confirmed))) + 
    geom_point() +
    ggtitle("Confirmed cases for each US State") +
    ylab("Country/Region") +
    xlab("Confirmed Cases")
```

Now we summarise the data based on the deaths by country
```{r DeathPerCountry}
report_03_11_2020 %>% 
  group_by(Country.Region) %>% 
  summarise(Deaths = sum(Deaths)) %>% 
  arrange(desc(Deaths))
```

Now we can plot the top 20 countries with the most COVID-19 related deaths
```{r deathChart}
report_03_11_2020 %>% 
  group_by(Country.Region) %>% 
  summarise(Deaths = sum(Deaths)) %>% 
  arrange(desc(Deaths)) %>% 
  slice(1:20) %>% 
  ggplot(aes(x = reorder(Country.Region, Deaths),  y = Deaths)) + 
    geom_bar(stat = 'identity') +
    ggtitle("The 20 countries with the most reported COV19-related deaths") +
    ylab("Country/Region") +
    xlab("Deaths")+
coord_flip()
```

Working with the time series data

```{r timeData}
time_series_confirmed <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")) %>%
  rename(Province.State = "Province/State", Country.Region = "Country/Region")
```
Checking data properties
```{r head2}
head(time_series_confirmed)
```
```{r convertToLong}
time_series_confirmed_long <- time_series_confirmed %>% 
               pivot_longer(-c(Province.State, Country.Region, Lat, Long),
                            names_to = "Date", values_to = "Confirmed") %>% 
               group_by(Country.Region,Date) %>% 
               summarise(Confirmed = sum(Confirmed))

# convert date to data format
time_series_confirmed_long$Date <- mdy(time_series_confirmed_long$Date)
# Thanks to Prof. Chris Sunderland for this code chunk
```

```{r headLong}
head(time_series_confirmed_long)
```
Now we can graph the confrimed US cases over time
```{r confirmedCasesUSTime}
time_series_confirmed_long %>% 
  filter (Country.Region == "US") %>% 
    ggplot(aes(x = Date,  y = Confirmed)) + 
    geom_point() +
    geom_line() +
    ggtitle("US Confirmed COVID-19 Cases")
```

We can also add in multiple countries
```{r confirmedCasesCountries}
time_series_confirmed_long %>% 
  filter (Country.Region %in% c("China","Japan", "Korea, South",
                                "Italy","Spain", "US")) %>% 
    ggplot(aes(x = Date,  y = Confirmed)) + 
    geom_point() +
    geom_line() +
    ggtitle("Confirmed COVID-19 Cases") +
    facet_wrap(~Country.Region, ncol=2, scales="free_y")
```
We can also put the multiple countries on a single graph
```{r confirmedcasessamegraph}
  time_series_confirmed_long %>% 
  filter (Country.Region %in% c("China","France","Italy", 
                                "Korea, South", "US")) %>% 
    ggplot(aes(x = Date,  y = Confirmed, color = Country.Region)) + 
    geom_point() +
    geom_line() +
    ggtitle("Confirmed COVID-19 Cases")
```

Visualizing data on maps
Here are two examples provided by Prof. Chris Sunderland and Anisa Dhana

```{r libraries 2}
library(maps)
library(viridis)
```
Get the world map
```{r world map}
world <- map_data("world")
```

Cutoffs based on the number of cases
```{r cutoffs}
mybreaks <- c(1, 20, 100, 1000, 50000)
```

Plot of data from March 11
```{r cases around world}
ggplot() +
  geom_polygon(data = world, aes(x=long, y = lat, group = group), fill="grey", alpha=0.3) +
  geom_point(data=time_series_confirmed, aes(x=Long, y=Lat, size=`2/25/20`, color=`2/25/20`),stroke=F, alpha=0.7) +
  scale_size_continuous(name="Cases", trans="log", range=c(1,7),breaks=mybreaks, labels = c("1-19", "20-99", "100-999", "1,000-49,999", "50,000+")) +
  # scale_alpha_continuous(name="Cases", trans="log", range=c(0.1, 0.9),breaks=mybreaks) +
  scale_color_viridis_c(option="inferno",name="Cases", trans="log",breaks=mybreaks, labels = c("1-19", "20-99", "100-999", "1,000-49,999", "50,000+")) +
  theme_void() + 
  guides( colour = guide_legend()) +
  labs(caption = "") +
  theme(
    legend.position = "bottom",
    text = element_text(color = "#22211d"),
    plot.background = element_rect(fill = "#ffffff", color = NA), 
    panel.background = element_rect(fill = "#ffffff", color = NA), 
    legend.background = element_rect(fill = "#ffffff", color = NA)
  )
```

Adaptation of code from Prof. Chris Sunderland
Get data, pivot to longer format and summarize

```{r}
time_series_confirmed_long2 <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")) %>%
    rename(Province.State = "Province/State", Country.Region = "Country/Region") %>%
    pivot_longer(-c(Province.State, Country.Region, Lat, Long),
    names_to = "Date", values_to = "cumulative_cases") %>%
    mutate(Date = mdy(Date) - days(1),
        Place = paste(Lat,Long,sep="_")) %>%
    group_by(Place,Date) %>%
        summarise(cumulative_cases = ifelse(sum(cumulative_cases)>0,
        sum(cumulative_cases),NA_real_),
        Lat = mean(Lat),
        Long = mean(Long)) %>%
    mutate(Pandemic_day = as.numeric(Date - min(Date)))
```
View the table
```{r head3}
head(time_series_confirmed_long2)
```
Plotting the data
```{r coivd-19Spread}
static <- ggplot(subset(time_series_confirmed_long2,Date %in% seq(min(Date),max(Date),7)),
            aes(x = Long, y = Lat, size = cumulative_cases/1000)) +
            borders("world", colour = NA, fill = "grey90") +
            theme_bw() +
            geom_point(shape = 21, color='purple', fill='purple', alpha = 0.5) +
            labs(title = 'COVID-19 spread',x = '', y = '',
                 size="Cases (x1000))") +
            theme(legend.position = "right") +
            coord_fixed(ratio=1)+
            facet_wrap(.~Date,nrow=3)
static
```

Exercises

Report Graphs:
```{r confirmedCasesChina}
tempData = subset(report_03_11_2020, Province.State=="Hubei")
values = rep(c(tempData$Confirmed, tempData$Deaths, tempData$Recovered))
condition = c("Confirmed", "Deaths", "Recovered")
tempData2 = data.frame(condition, values)
ggplot(tempData2, aes(x=condition, y=values)) + geom_bar(stat = 'identity') + ylab("Number of Reported Cases") + xlab("Status") + ggtitle("State of COVID19 Cases in Hubei Province 3/11")
```

```{r US Recovered Cases}
report_03_11_2020 %>%
  filter (Country.Region == "US") %>%
  group_by(Province.State) %>% 
  summarise(Recovered = sum(Recovered)) %>% 
  arrange(desc(Recovered)) %>% 
  slice(1:5) %>%
  ggplot(aes(x = reorder(Province.State, Recovered),  y = Recovered)) + 
    geom_bar(stat = 'identity') +
    ggtitle("The 5 states with the highest amount of recovered cases") +
    ylab("State") +
    xlab("Recovered Cases")
  
```

Time Series Graphs:
```{r European Union}
time_series_confirmed_long %>% 
  filter (Country.Region %in% c("France","Spain","Sweden", 
                                "Germany", "Finland", "Poland", "Italy", "Romania","Greece", "Bulgaria", "Hungary")) %>% 
    ggplot(aes(x = Date,  y = Confirmed, color = Country.Region)) + 
    geom_point() +
    geom_line() +
    ggtitle("Confirmed COVID-19 Cases Across 10 Largest EU Countries")
```

```{r AustraliaRegions}
time_series_confirmed_long3 <- time_series_confirmed %>% 
               pivot_longer(-c(Province.State, Country.Region, Lat, Long),
                            names_to = "Date", values_to = "Confirmed")

time_series_confirmed_long3$Date <- mdy(time_series_confirmed_long3$Date)

time_series_confirmed_long3 %>% 
  filter (Country.Region %in% c("Australia")) %>%
  group_by(Province.State) %>%
    ggplot(aes(x = Date,  y = Confirmed,  color = Province.State)) + 
    geom_point() +
    geom_line() +
    ggtitle("Confirmed COVID-19 Cases Across Australian Provinces") +
    facet_wrap(~Country.Region, ncol=2, scales="free_y")
          
```

Map Graphs

Total confirmed deaths gloablly (Outdated)
```{r Total Deaths Gloablly}
time_series_death <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv"))

ggplot() +
  geom_polygon(data = world, aes(x=long, y = lat, group = group), fill="grey", alpha=0.3) +
  geom_point(data=time_series_death, aes(x=Long, y=Lat, size=`2/25/20`, color=`2/25/20`),stroke=F, alpha=0.7) +
  scale_size_continuous(name="Cases", trans="log", range=c(1,7),breaks=mybreaks, labels = c("1-19", "20-99", "100-999", "1,000-3,500", "4,000+")) +
  # scale_alpha_continuous(name="Cases", trans="log", range=c(0.1, 0.9),breaks=mybreaks) +
  scale_color_viridis_c(option="inferno",name="Cases", trans="log",breaks=mybreaks, labels = c("1-19", "20-99", "100-999", "1,000-3,500", "4,000+")) +
  theme_void() + 
  guides( colour = guide_legend()) +
  labs(caption = "") +
  theme(
    legend.position = "bottom",
    text = element_text(color = "#22211d"),
    plot.background = element_rect(fill = "#ffffff", color = NA), 
    panel.background = element_rect(fill = "#ffffff", color = NA), 
    legend.background = element_rect(fill = "#ffffff", color = NA)
  )
```


```{r }
time_series_death_long <- read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv")) %>%
    rename(Province.State = "Province/State", Country.Region = "Country/Region") %>%
    filter(Country.Region == "Italy" | Country.Region == "China") %>%
    pivot_longer(-c(Province.State, Country.Region, Lat, Long),
    names_to = "Date", values_to = "cumulative_cases") %>%
    mutate(Date = mdy(Date) - days(1),
        Place = paste(Lat,Long,sep="_")) %>%
    group_by(Place,Date) %>%
        summarise(cumulative_cases = ifelse(sum(cumulative_cases)>0,
        sum(cumulative_cases),NA_real_),
        Lat = mean(Lat),
        Long = mean(Long)) %>%
    mutate(Pandemic_day = as.numeric(Date - min(Date)))

static2 <- ggplot(subset(time_series_death_long,Date %in% seq(min(Date),max(Date),7)),
            aes(x = Long, y = Lat, size = cumulative_cases)) +
            borders("world", colour = NA, fill = "grey90") +
            theme_bw() +
            geom_point(shape = 21, color='red', fill='red', alpha = 0.5) +
            labs(title = 'Death Toll in Italy & China',x = '', y = '',
                 size="Deaths") +
            theme(legend.position = "right") +
            coord_fixed(ratio=1)+
            facet_wrap(.~Date,nrow=3)

static2
```




